% !TEX root = creche.tex
\documentclass[creche.tex]{subfiles}
\chapter{Topological dynamics}\label{dynamics}


\def\medrel#1{\parbox[t]{6ex}{$\displaystyle\hfil #1$}}
\def\ceq#1#2#3{\parbox[t]{18ex}{$\displaystyle #1$}\medrel{#2}{$\displaystyle #3$}}

Scratch paper

\section{}

\def\mrX{{\mr\X}}
Let $G$ be a semigroup that acts from the left on some set $\mrX\subseteq \U^{|{\mr x}|}$. The notion below are relative to $G$ and $\mrX$, though this will not be displayed in the notation.

If $\mrD\subseteq\mrX$ we write $g^{-1}\mrD$ for the set $\{{\mr x}: g{\mr x}\in\mrD\}$. Clearly, if $\mrD$ is the set defined by $\phi({\mr x})$, then $g^{-1}\,\mrD$ is the set defined by $\phi(g\,{\mr x})$. 

We say that a set $\mrD\subseteq \mrX$ is \emph{quasi-invariant\/} if for any finitely many $g_1,\dots,g_n\in G$ 

\ceq{\hfill\bigcap^n_{i=1}g_i^{-1}\mrD}{\neq}{\0} 

We say that a formula $\phi({\mr x})$ is \emph{quasi-invariant\/} if the set $\phi(\mrU)$ is such. A type $p({\mr x})$ is \emph{quasi-invariant\/} if so are all formulas in $p({\mr x})$. 

We say that $\mrD$ is \emph{strongly quasi-invariant\/} if for every formula $\phi({\mr x})\in L(\U)$ at least one of $\phi({\mr\U})\cap\mrD$ or  $\neg\phi({\mr\U})\cap\mrD$ is quasi-invariant.

We say that $\mrD$ is \emph{strongly quasi-invariant\/} if for every finite number of formulas $\phi_i({\mr x})\in L(\U)$ that cover $\mrD$, at least one of the sets $\phi_i(\mrU)\cap\mrD$ is quasi-invariant.

\begin{lemma}
If $\mrD$ is strongly quasi-invariant then for every $\phi({\mr x})\in L(\U)$ at least one of $\phi({\mr\U})\cap\mrD$ or $\neg\phi({\mr\U})\cap\mrD$ is strongly quasi-invariant.
\end{lemma}


\begin{proof}


\end{proof}

\begin{corollary}
Every strongly quasi-invariant type extends to a global strongly quasi-invariant type.
\end{corollary}

\begin{proof}


\end{proof}



\section{Hindman theorem                            }


\def\fp{{\rm fp}}

Let $G$ be a semigroup. If $\bar a=\<a_i:i<\alpha\>$ is a tuple of elements of $G$ we write $\fp(\bar a)$ for the sets of finite products (in increasing order) of elements of $\bar a$. Namely,

\ceq{\hfill\emph{$\fp(\bar a)$}}{=}{\Big\{a_{i_1}\cdot\dots\cdot a_{i_n}\ :\quad i_0<\dots<i_{n-1}<\alpha,\ \  n\in\omega\Big\}}.

We agree that when $\bar a$ is the empty tuple then $\fp(\bar a)=\0$.

\begin{Hindman}
Let $G$ be an infinite semigroup. Then for every finite coloring of $G$ there is an infinite tuple $\bar c$ such that $\fp(\bar c)$ is monochromatic.
\end{Hindman}

When $G$ is commutative this amount to claiming that there is an infinite monochromatic set cosed under finite sum of distinct elements.

\begin{proof}
We follow closely the proof of Ramsey's Theorem~\ref{thm_Ramsey}. 
Let $L$ be a language that contains the semigroup operation \emph{$\cdot$\/} and the unary relation symbols $r_0,\dots,r_{k-1}$ that correspond to the coloring of $G$. 

We may assume that $G$ is an elementary substructure of some large saturated model $\U$. Pick any type $p({\mr x})\in S(\U)$ finitely satisfied in $G$ but not realized in $G$ and let ${\mr\bar c}=\<{\mr c_i}:i<\omega\>$ be a coheir sequence of $p({\mr x})$.

Suppose not and let $\phi({\mr x})\in p$ be such that $a\cdot b\notmodels\phi({\mr x})$. Then there is a $g\in G$ such that $a\cdot g\models\neg\phi(x)$. Then $\neg\phi(x\cdot g)\in p$. Which contradicts the invariance of $p(x)$.






which we assume to be a structure in a language $L$ that contains the semigroup operation and a unary relation for every subset of $G$. If $X\subseteq G$ we write $\xi_X(x)$ for the corresponding predicate. We assume that $\U$ is a large saturated elementary extension of $G$. For $\phi(x)\in L(\U)$ we write $\phi(G)$ for $G\cap\phi(\U)$.
\end{proof}


Let $q(x)\subseteq L(G)$ be quasi-invariant. Let $p(x)\in S(\U)$ be a global extension of $q(x)$ that is finitely satisfied in $G$. Then $p(x)$ is quasi invariant. Suppose not. Let $\phi(x\,;a)$ be a formula in $p(x)$ such that 

\ceq{\hfill\neg\E x \bigwedge^n_{i=1}\phi(x\,g_i\,; a)}{}{} 

We claim that  

\ssf{\#}\quad $b\,a\models p_{\restriction G}(x)$ for every $a\models p_{\restriction G}(x)$ and $b\models p_{\restriction G,a}(x)$.

Suppose not then $\neg\phi(b\,a)$ for some formula $\phi(x)\in p$. From $b\models p_{\restriction G,a}(x)$ we obtain $\neg\phi(g\,a)$ for some $g\in G$. Hence $\neg\phi(g\,x)\in p$.

Let $p(x)\subseteq L(\U)$ be maximal among the quasi-invariant types that are finitely satisfied in $G$. We claim that $p(x)$ is complete. Suppose for a contradiction that there is $\phi(x)\in L(\U)$ such that both $\phi(x)$ and $\neg\phi(x)$ are consistent with $p(x)$. Without loss of generality we can assume that $\phi(x)$ is not quasi-invariant and that $\neg\phi(x)$ is not finitely satisfied in $G$. Then there is a $\psi(x)\in p$ such that $\psi(G)\subseteq \phi(G)$ and 

\ceq{\hfill\neg\E x}{}{ \bigwedge^n_{i=1}\psi(g_i\,x)\wedge\phi(g_i\,x)} 




\section{Hindman's theorem                                }
\def\ext{{\rm ext}}
This section is dedicated to another application of coheirs. The argument is essentially the same as for Ramsey's theorem but we apply it to a special coheir that we obtain from Lemma~\ref{}. This lemma relies on ideas taken from topological dynamics which will be discussed in more depth in Chapter~\ref{dynamics}.

Let $G$ be an infinite semigroup which we assume to be a structure in a language $L$ that extends the language of multiplicative groups. The set of externally definable subsets of $G$ is

\ceq{\hfill\ext(G)}{=}{\Big\{\phi(G)\ :\ \phi(x)\in L(\U)\Big\}}

where $\phi(G)$ stand for $\phi(\U)\cap G$, the trace of $\phi(\U)$ on $G$. 

There is an obvious right-action of $\U$ on its subsets, namely $\D\mapsto \D{\cdot}a$. In particular, the action takes definable sets to definable sets. Namely, the set defined by $\phi(x)$ is mapped to a the set defined by 

\ceq{\hfill\emph{$\phi_{\cdot a^{-1}}(x)$}}{=}{\E y\big[\phi(y)\wedge x{=}y{\cdot}a\big]}. 

By taking the trace on $G$ of the definable sets, we obtain an action of $\U$ on $\ext(G)$. Namely $\phi(G)$ maps to $\phi_{\cdot a^{-1}}(G)$ defined above. 

This action is continuous in the $G$-topology on $\U$. In particular, it is an action on $\U/{\equiv_G}$.



with a unary relation for every subset of $G$. If $X\subseteq G$ we write $\xi_X(x)$ for the corresponding predicate. We assume that $\U$ is a large saturated elementary extension of $G$. For $\phi(x)\in L(\U)$ we write $\phi(G)$ for $G\cap\phi(\U)$.





\begin{lemma}
There is a $g\in G$ such that $a\equiv$.
\end{lemma}

\begin{proof}

\end{proof}

The motivation for the choice of the language is to obtain the following: every type in $p(x)\in S(G)$ has a unique global coheir. In fact, as the formula $\phi(x)\niff\xi_{\phi(G)}(x)$ is not satisfied in $G$, all global types that are finitely satisfied in $G$ have to contain the formula $\phi(x)\iff\xi_{\phi(G)}(x)$. Therefore the type 

\ceq{\hfill p'(x)}{=}{\big\{\phi(x)\ :\  \xi_{\phi(G)}(x)\in p\big\}}

is the only possible global coheir of $p(x)$.

We want to extend the group operation of $G$ to a multiplication $*$ on $\U/{\equiv_G}$. (The group multiplication in $\U$ is never compatible with the relation $\equiv_G$, see Exercise~\ref{ex_group_operation}.) Given $a,b\in\U$ let $p(x)$ be the unique global coheir of $\tp(\dot a/G)$ and let $\dot a\models p_{\restriction G,b}(x)$ and let $c=\dot a\cdot b$. Note that $\tp(c/G)$ only depends on $\tp(a/G)$ and $\tp(b/G)$ in particular it does not depend on the choice of $\dot a$.

\begin{lemma}
There exists a type $p(x)\in S(\U)$ that is finitely satisfied in $G$ and such that 

\ssf{\#}\quad $a\cdot b\models p_{\restriction G}(x)$ for every $a\models p_{\restriction G}(x)$ and $b\models p_{\restriction G,a}$. 
\end{lemma}

\begin{proof}
Let $q(x)\subseteq L(\U)$ be the maximal type for which \ssf{\#} holds. We claim that for every $q(a\cdot x)\iff q(x)$. 
\end{proof}

Let $\rho(a,b,c)$ be the relation that holds if $c=\dot a\cdot b$ for some $\dot a\models p_{\restriction G,b}(x)$, where $p(x)$ is a global coheir of $\tp(\dot a/G)$.


\begin{proposition}
The relation $\rho(a,b,c)$ is invariant over $G$.
\end{proposition}

\begin{proof}
Fix $\dot a$ as above such that $c=\dot a\cdot b$ and let $f\in\Aut(\U/G)$. As $f(\dot a)\models p_{\restriction G,f(b)}(x)$, ...
\end{proof}

By the $G$-invariance of $p(x)$, the type of $a'\cdot b$

\def\fp{{\rm fp}}

If $\bar a=\<a_i:i<\alpha\>$ is a tuple of elements of $\U$ we write $\fp(\bar a)$ for the sets of finite products (in increasing order) of elements of $\bar a$. Namely,

\ceq{\hfill\emph{$\fp(\bar a)$}}{=}{\Big\{a_{i_1}\cdot\dots\cdot a_{i_n}\ :\quad i_0<\dots<i_{n-1}<\alpha,\ \  n\in\omega\Big\}}.

We agree that when $\bar a$ is the empty tuple then $\fp(\bar a)=\0$.

\begin{theorem}
Let $G$ be an infinite semigroup. Then for every finite coloring of $G$ there is an infinite tuple $\bar c$ such that $\fp(\bar c)$ is monochromatic.
\end{theorem}

When $G$ is commutative this amount to claiming that there is an infinite monochromatic set cosed under finite sum of distinct elements.

\begin{proof}
We follow quite closely the proof of Ramsey's Theorem~\ref{thm_Ramsey}. We may assume that the colors corresponds to definable sets that form a partition of $G$ and, by elementarity, of $\U$. Let $p(x)\in S(\U)$ be the global type obtained from Lemma~\ref{}. For all $i<\omega$ define

\ceq{\hfill c_{i+1}}{\models}{p_{\restriction G,\,\fp(c_{\restriction i})}}

By  Lemma~\ref{}, the set $\fp(\bar c)$ is monochromatic. It need not be contained in $G$ though. We use $\bar c$ to define $\bar a\in G^\omega$ such that $\fp(\bar a)$ is monochromatic.

Assume as induction hypothesis that $\fp({\mr a_{\restriction i}},{\mr c_{\restriction n}})$ is monochromatic. Our goal is to find ${\mr a_i}$ such as the same property holds for ${\mr a_{\restriction i}},{\mr a_i},{\mr c_{\restriction n}}$. By the indiscernibility of ${\mr\bar c}$, the property holds for  ${\mr a_{\restriction i}},{\mr c_{\restriction n}},{\mr{c_n}}$.  And this can be written by a formula $\phi({\mr a_{\restriction i}},{\mr c_{\restriction n}},{\mr{c_n}})$. As ${\mr\bar c}$ is a coheir sequence, by Lemma~\ref{lem_coheir_property} we can find  ${\mr a_i}\in M$ such that $\phi({\mr a_{\restriction i}},{\mr c_{\restriction n}},{\mr a_i})$. So, as the order is irrelevant,  ${\mr a_{\restriction i}},{\mr a_i},{\mr c_{\restriction n}}$ satisfies the induction hypothesis.
\end{proof}

First note that here is natural action of $\U$ on the subsets of $\U$. Namely, for $a\in\U$ and $\D\subseteq\U$ 

\ceq{\hfill \D}{\stackrel{a}{\mapsto}}{a{\cdot}\D=\{a\,b\ :\ b\in\D\}.}

This induces an action of $\U$ on $\P(G)$, namely,

\ceq{\hfill X}{\stackrel{a}{\mapsto}}{G\ \cap\ a{\cdot}\phi_X(\U).}

It is immediate, but crucial, to note that this action only depends on $\tp(a/G)$. 

Let $b\in\U$ be We define $a*b$ as 


\begin{exercise}\label{ex_group_operation}
Let $G$ be such that ..... Prove that there are $a\equiv_G a'$ and $b\equiv_G b'$ such that $a\,b\equiv_G a'b'$. The operation we define below will only be a subgroup multiplication.\QED
\end{exercise}
\def\Ex{\Ee\kern-.2ex\mbox{$\scriptscriptstyle\X$}}



\section{Altra prova                               }

Let $G$ be an infinite semigroup which we assume to be a structure in a language $L$ comprising a symbol for multiplication and possibly other symbols. 


In this section we write 

\ceq{\hfill\emph{$a\nonforkc_Ab$}}{\IFF}{\phi(a\,;b) \textrm{ for all }\phi(z\,;x)\in L(A)\textrm{ such that }A\subseteq\phi(a\,;\U)}

\ceq{}{\IFF}{\phi(a\,;A)\neq\0 \textrm{ for all }\phi(z\,;x)\in L(A)\textrm{ such that }\phi(a\,;b)}

which reads, $\tp(b/A,a)$ is finitely satisfied in $A$. We may write \emph{$b\cnonfork_Aa$} for $a\nonforkc_Ab$.

\ceq{\hfill\emph{$a\nonforkc_Ax$}}{=}{\Big\{\phi(a\,;x)\ :\ \phi(z\,;x)\in L(A)\textrm{ and }A\subseteq\phi(a\,;\U)\Big\}}

We shall use, sometimes without reference, the following easy lemma

\begin{lemma}
If $a\nonforkc_Ac$ then for every $b$ there is $c'$ such that $a,b\nonforkc_Ac'\equiv_{A,a}c$. 
\end{lemma}
\begin{proof}
Let $p(x)$ be a global coheir of $\tp(c/A,a)$, which exists by Proposition~\ref{prop_exisntence_coheirs}. Then any $c'\models p_{\restriction A,a,b}(x)$ proves the lemma.  
\end{proof}

\begin{lemma}
Suppose that for every $a\in\U$ the type $a\nonforkc_Gx$ is complete. Let $q(x)\subseteq L(G)$ be a maximal type such that $a{\cdot}b\models q(x)$ for every $a,b\models q(x)$ such that $a\nonforkc_Gb$. Then $q(x)$ is a complete type.
\end{lemma}

\begin{proof}
Fix an arbitrary $a\models q(x)$. We prove that

\ceq{\ssf{1}\hfill q(x)}{\imp}{\E x'\ \big[x\equiv_Ga{\cdot}x'\ \wedge\ a\nonforkc_Gx'\big]}

Note that, by $G$-invariance, the r.h.s.\@ is equivalent to a type over $G$. Therefore, by the maximality of $q(x)$, it suffices to show that the r.h.s.\@ is consistent with $q(x)$ and closed under multiplication. As $q(x)$ is closed under multiplication, consistency is witnessed by $a{\cdot}a'$ where $a'$ is any element such that $a\nonforkc_Ga'\models q(x)$.
%
%Now, let $b\nonforkc_Gc$ be two realizations of the r.h.s.\@ and let $b',c'$ witness  the respective existential quantifiers. Note that, by Lemma~\ref{lem_coheirext}, we can choose $c'$ such that $a,b,b'\nonforkc_Gc'$. Therefore, from $b\equiv_Ga{\cdot}b'$ we obtain  $b{\cdot}c'\equiv_Ga{\cdot}b'{\cdot}c'$. From transitivity we obtain $a\nonforkc_Gb'{\cdot}c'$. Then $b'{\cdot}c'$ witnesses that $b{\cdot}c'$ satisfies the r.h.s. By elementarity, also $b{\cdot}c$  satisfies the r.h.s.

Now, let $b\nonforkc_Gc$ be two realizations of the r.h.s.\@ and let $b'$ witness the existential quantifier for $b$. By Lemma~\ref{lem_coheirext}, there is $c'$ such that $a, b, b'\nonforkc_Gc'\equiv_{G,b}c$. Then, from $b\equiv_Ga{\cdot}b'$ we obtain  $b{\cdot}c'\equiv_Ga{\cdot}b'{\cdot}c'$.  By transitivity $a\nonforkc_Gb'{\cdot}c'$. Then $b'{\cdot}c'$ witnesses that $b{\cdot}c'$ satisfies the r.h.s. By elementarity, also $b{\cdot}c$  satisfies the r.h.s. This proves \ssf{1}.


We now claim that 

\ceq{\ssf{2}\hfill q(x)}{\imp}{\E x'\ \big[a\equiv_Ga{\cdot}x'\ \wedge\ a\nonforkc_Gx'\equiv_Gx\big]}

We proceed as for the proof of \ssf{1}. Consistency is again witnessed by $a$, as can be easily verified substituting $a$ for $x$ in \ssf{1}. Now, let $b\nonforkc_Gc$ be two realizations of the r.h.s.\@ and let $b',c'$ witness the respective existential quantifiers. By Lemma~\ref{lem_coheirext}, we can also require that $a,b,b'\nonforkc c'\equiv_{G}c$. Then from $a\equiv_Ga{\cdot}b'$ we obtain $a{\cdot}c'\equiv_Ga{\cdot}b'{\cdot}c'$ hence $a\equiv_Ga{\cdot}b'{\cdot}c'$. From the completeness of $b\nonforkc_G x$ we infer $c\equiv_{Gb}c'$. Hence $b{\cdot}c\equiv_G b{\cdot}c'$. As $b,b'\nonforkc_Gc'$, we obtain that $b'{\cdot}c'$ witnesses that $b{\cdot}c$ satisfies the r.h.s. of \ssf{2}.

Finally we prove that

\ceq{\ssf{3}\hfill q(x)}{\imp}{a\equiv_Gx}

As for \ssf{1} and \ssf{2}, it suffices to prove that if $b\nonforkc_Gc$ are two realizations of the r.h.s.\@ then so is also $b{\cdot}c$. We can assume $a=b$. As $c\models q(x)$, from \ssf{2} we obtain $c'$ such that $a\equiv_Ga{\cdot}c'$ and $a\nonforkc_Gc'\equiv_Gc$. As $a\nonforkc_Gc$ we can assume without loss of generality that $c'\equiv_{G,a}c$ so $a\equiv_Ga{\cdot}c$ follows.

As $a\models q(x)$, \ssf{3} implies that $q(x)=\tp(a/G)$. 
\end{proof}

We identity each element $a\in\U$ with its left action on $\U$, i.e.\@ with the function $a(x)=a{\cdot} x$. We write \emph{$a^{-1}\D$} for $\big\{b:a{\cdot} b\in\D\big\}$, i.e.\@ the inverse image of $\D$ under the action of $a$. If $\D$ is definable by the formula $\phi(x)$ then $a^{-1}\D$ is definable by the formula $\phi(a{\cdot} x)$. This action on $L_x(\U)$ induces in the obvious way an action on $S_x(\U)$, namely $p(x)$ maps to $p(a{\cdot} x)$.

\begin{assumption} 
For every $G$-definable set $\D\subseteq\U$ and every element $a\in\U$ there is a unique $G$-definable set $\C$ such that $\C=_Ga^{-1}\D$.\QED
\end{assumption}

The assumption above allows to define an action of $\U$ on $L_x(G)$. For each formula $\phi(x)\in L(G)$ we define \emph{${}^a\kern-1.2pt \phi(x)$\/} to be the unique (up to equivalence) formula in $L(G)$ that is equivalent to $\phi(a{\cdot} x)$ for all $x\in G$. Given $p(x)\in S(G)$ we define \emph{${}^a\kern-.8pt p(x)$\/}$\in S(G)$ to be the type that contains the formulas ${}^a\kern-1.2pt \phi(x)$ for all $\phi(x)\in p$. 

Note that the actions defined above factorize through $\equiv_G$. So, $\U/{\equiv_G}$ acts on $S_x(G)$. As these two sets can be identified, we actually have an operation on $S_x(G)$. For $p,q\in S_x(G)$ we define \emph{$p*q$\/} to be the type ${}^a\kern-.8pt q(x)$ for some (any) $a\models p(x)$.

\begin{lemma}
For every $p,q\in S_x(G)$, every $a\models p(x)$, and every $b\models q(x)$ that is ??,  we have $a\cdot b\models (p*q) (x)$.
\end{lemma}

\begin{proof}
Let ${}^a\kern-1.2pt \phi(x)\in {}^a\kern-.8pt q$ be arbitrary. Then $\phi(a{\cdot} x)\iff{}^a\kern-1.2pt \phi(x)$ holds for every $x\in G$. Then it holds for $b$ and the lemma follows.
\end{proof}

By the lemma we may consistently define $p*q$ to be $\tp(a{\cdot} b/G)$ for some (any) $a\models p(x)$ and $b\models {}^a\kern-.8pt p$.

ttlews


We write $a\phi(x)$ for the (any) formula over $G$ that defines $a^{-1}\D$. This is well-defined up to equivalence.


Above we have define a right action on $G$-definable sets and a left action on formulas in $L_x(G)$. The latter induces anaction on . Note that these actions factorize through $\equiv_G$. 

Now we define a semigroup operation on $\U/{\equiv_G}$.  



There is a natural left action of $\U$ on $L_x(\U)$-formulas. Namely, the action of $a\in\U$ maps $\phi(x)$ to the formula $\phi(a{\cdot} x)$ which here we denote by \emph{$a^{-1}\phi(x)$\/}. The notation is justified by the fact that the set defined by $a^{-1}\phi(x)$ is the inverse image of $\phi(\U)$ under the action of $a$. The same notation is also applied to types, we write \emph{$a^{-1}p(x)$\/} for $p(a{\cdot} x)$.

The externally definable subsets of $G$ are the traces on $G$ of the definable sets. The set of all these traces is denoted by 

\ceq{\hfill\emph{\ext(G)}}{=}{\Big\{\phi(\U)\cap G\ :\ \phi(x)\in L(\U)\Big\}}

The action of $\U$ on on formulas induces an action on $\ext(G)$. Clearly, this action factorize through $\equiv_G$.

We now assume that all sets in $\ext(G)$ are the trace of a $G$-definable set and that this set is unique.

If $G$ is a model then by elementarity 

The action of $\U/{equiv_G}$ on $\ext(G)$ induces an action on $L_x(G)$. The action of $a$ on the definable set $\D$ maps it to the unique definable set $\C =_G a^{-1}\D$.



The action above applies to all formulas in $L_x(\U)$ but here we are mostly interested in the trace on $G$ of the definable sets. 


formulas $L_x(G)$ and types in $S_x(G)$. If $p\in S_x(G)$ and $a\equiv_G b$ then 






There is an obvious right-action of $\U$ on its subsets, namely $\D\mapsto \D{\cdot}a$. In particular, the action takes definable sets to definable sets: the set defined by $\phi(x)$ is mapped to a the set defined by 

\ceq{\hfill a^{-1}\phi(x)}{=}{\E y\big[\phi(y)\wedge x{=}a{\cdot}y\big]}. 

We stress that $a^{-1}\phi(x)$ is just an abbreviation. If the semigroup $G$ happens to be a group, then  $a^{-1}\phi(\U)=\{a^{-1}b:b\in\phi(\U)\}$. By taking the trace on $G$ of the definable sets, we obtain an action of $\U$ on $\ext(G)$. Namely $\phi(x)$ maps to $\phi_{a^{-1}}(x)$ defined above. 

A 


This action is continuous in the $G$-topology on $\U$. In particular, it is an action on $\U/{\equiv_G}$.


\section{Altra prova (gruppi, non semigruppi)                             }


Let $G$ be an infinite group which we assume to be a structure in a language $L$ that contains the language of multiplicative groups and possibly other symbols.

For $a\in\U$ and $\D\subseteq\U$, we write $a\D$ for the set $\{ab:b\in\D\}$. If $\D$ is defined by the formula $\phi(x)$ then $a\D$ is defined by the formula $\phi(a^{-1}{\cdot}\, x)$.

\begin{assumption} 
For every $G$-definable set $\D\subseteq\U$ and every $a\in\U$ there is a $G$-definable set $\C$ such that $\C=_Ga\D$.\QED
\end{assumption}

The assumption above allows to define an action of $\U$ on $L_x(G)$. For each formula $\phi(x)\in L(G)$ we define \emph{$\phi^a(x)$\/} to be the unique (up to equivalence) formula in $L(G)$ that is equivalent to $\phi(a^{-1}{\cdot}\, x)$ for all $x\in G$. Given $p(x)\in S(G)$ we define \emph{$p^a(x)$\/}$\in S(G)$ to be the type that contains the formulas $\phi^a(x)$ for all $\phi(x)\in p$.

Note that the actions defined above factorize through $\equiv_G$. So, $\U/{\equiv_G}$ acts on $S_x(G)$. As these two sets can be identified, we actually have an operation on $S_x(G)$. For $p,q\in S_x(G)$ we define \emph{$p*q$\/} to be the type $q^a(x)$ for some (any) $a\models p(x)$.

\begin{lemma}
For every $p,q\in S_x(G)$, every $a\models p(x)$, and every $b\models q(x)$ such that $a\nonforkc_G b$ we have $a\cdot b\models (p*q) (x)$.
\end{lemma}

\begin{proof}
Let $\phi(x)\in p*q$ be arbitrary. Then $\phi(x)\in q^a$, which we may equivalintly write as $\phi^{a^{-1}}(x)\in q$. Hence, by assumption, $b\models \phi(a{\cdot} x)$. As $\phi(a{\cdot} x)\iff\phi^{a^{-1}}(x)$ holds for every $x\in G$, by independence it holds also for $b$. Therefore $\phi(a{\cdot} b)$ and the lemma follows.
\end{proof}

\section{Altra prova (semigruppi supersemplificta)                 }

Let $G$ be an infinite semigroup which we assume to be a structure in a language $L$ that has a binary function symbol for multiplication and possibly other symbols. 

\begin{lemma}
Let $q(x)\subseteq L(G)$ be a maximal type that is 
\begin{itemize}
\item[1.] finitely satisfied in $G$
\item[2.] such that $a{\cdot} b\models q(x)$ for every $a\models q(x)$ and $b\nonforkc_G a$.
\end{itemize}
Then $q(x)$ is complete.
\end{lemma}

\begin{proof}
Let $p(x)$ be a global coheir of $q(x)$. Let $\<a_i:i\in\omega\>$ be a coheir sequence of $p(x)$. Let $c_i=a_0\cdot\dots\cdot a_i$. By assumption $c_i\models q(x)$. Let $p_i(x)=\tp(c_i/G)$. Define 

\ceq{\hfill p(x)}{=}{\bigcap_{i\in\omega}p_i(x).}

We claim that $p(x)=q(x)$. It suffices to prove that $p(x)$ is closed under  
\end{proof}

