% !TEX root = creche.tex
\documentclass[creche.tex]{subfiles}
\begin{document}
\chapter{Non-independence}
\label{nip}

\def\medrel#1{\parbox[t]{6ex}{$\displaystyle\hfil #1$}}
\def\ceq#1#2#3{\parbox{25ex}{$\displaystyle #1$}\medrel{#2}$\displaystyle  #3$}



In this chapter we fix a signature $L$, a complete theory $T$ without finite models, and a saturated model $\U$ of inaccessible cardinality $\kappa$ strictly larger than $|L|$.
The notation and implicit assumptions are as in Section~\hyperref[monster]{\ref*{saturation}.\ref*{monster}}.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Vaknik-Chevronenkis dimension}
\label{approximable}

\def\ceq#1#2#3{\parbox{25ex}{$\displaystyle #1$}\medrel{#2}$\displaystyle  #3$}

We say that the formula $\phi({\mr x}\,;{\gr z})\in L$ has \emph{Vaknik-Chevronenkis dimension $n$\/} if this is the largest finite cardinatity of a set ${\mr A}\subseteq\U^{|{\mr x}|}$ such that $|S_\phi({\mr A})|=2^n$. If no finite $n$ exists, we say that it has \emph{infinite VC-dimension}. 

Note that the condition $|S_\phi({\mr A})|=2^n$ is equivalent to saying that every subset of ${\mr A}$ is the trace of some definable set of sort $\phi({\mr x}\,;{\gr z})$.

Arguing by compactness we obtain the following proposition whose proof is left as an exercise for the reader.

\begin{proposition}
The following are equivalent 
\begin{itemize}
\item[1.] $\phi({\mr x}\,;{\gr z})\in L$ has finite VC-dimension;
\item[2.] there is no infinite set ${\mr A}\subseteq\U^{|{\mr x}|}$ such that  every subset of ${\mr A}$ is the trace of some definable set of sort $\phi({\mr x}\,;{\gr z})$.\QED
\end{itemize}
\end{proposition}

We say that the a sequence of sentences $\<\phi_i:i<\omega\>$ converges if the truth value of $\phi_i$ is eventually constant.  

\begin{lemma}\label{lem_altrank}
The following are equivalent
\begin{itemize}
\item[1.] $\phi({\mr x}\,;{\gr z})\in L$ has finite Vaknik-Chevronenkis dimension;
\item[2.] $\<\phi({\mr a_i}\,;{\gr b}):i<\omega\>$ converges for any ${\gr b}$ and any indiscernible sequence $\<{\mr a_i}:i<\omega\>$.
\end{itemize}
\end{lemma}

\begin{proof}
\ssf{1}$\IMP$\ssf{2} Assume $\neg$\ssf{2}. It suffices to prove that for every $I\subseteq\omega$ the type that says 

\ceq{\hfill\phi({\mr a_i}\,;{\gr z})}{\IFF}{i\in I}

is finitely consistent. Let $\psi_{I\restriction n}({\mr x_0},\dots,{\mr x_{n-1}})$ the formula that says that there is a ${\gr z}$ such that

\ceq{\hfill\phi({\mr x_i}\,;{\gr z})}{\IFF}{i\in I}\quad for all $i<n$.

is finitely consistent we need to prove that $\psi_{I\restriction n}({\mr a_0},\dots,{\mr a_{n-1}})$ is true. If there is a ${\gr b}$ such that the truth value of $\<\phi({\mr a_i}\,;{\gr b}):i<\omega\>$ oscillates at least $n$ times, then we can find $k_0<\dots<k_{n-1}$ such that 


\ceq{\hfill \phi({\mr a_{k_i}}\,;{\gr b})}{\IFF}{i\in I.}

This proves that $\psi_{I\restriction n}({\mr a_{k_0}},\dots,{\mr a_{k_{n-1}}})$ holds true. Therefore, by indiscernibility, $\psi_{I\restriction n}({\mr a_0},\dots,{\mr a_{n-1}})$ holds true.

\ssf{2}$\IMP$\ssf{1} Let $I\subseteq\omega$ be the set of even integers. Assume $\neg$\ssf{1} and let  $\<c_i:i<\omega\>$ be an infinite sequence that is shattered by $\phi({\mr x}\,;{\gr z})$.
Let  $\<{\mr a_i}:i<\omega\>$ be an indiscernible sequence with the same EM-type as $\<c_i:i<\omega\>$. Then $\<a_i:i<\omega\>$ satisfies $\psi_{I\restriction n}({\mr x_0},\dots,{\mr x_{n-1}})$ for all $n$. By compactness there is a $b$ such that

\ceq{\hfill\phi({\mr a_i}\,;{\gr z})}{\IFF}{i}\quad is even.

This proves $\neg$\ssf{2}.
\end{proof}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Honest definitions}
\label{nip-theory}

If all formulas have a finite VC-dimension, we say that the theory $T$ has the \emph{non-independence property},for short, $T$ is \emph{nip}.

We write \emph{$\neg^n$\/} for $\neg\!\stackrel{n\ \rm times}{\dots\dots\dots}\!\neg$. We abbreviate $\neg^n(\cdot\in\cdot)$ as \emph{$\notin^n$}.

\begin{lemma}
Let $\grD$ be a set approximable by a formula with finite VC-dimension. Then every global $A$-invariant type $p({\gr z})$ contains a formula $\psi({\gr z})$ such that either $\psi({\gr z})\imp {\gr z}\in\grD$ or $\psi({\gr z})\imp {\gr z}\notin\grD$.
\end{lemma}

\begin{proof}
By Lemma~\ref{lem_altrank}, there is no infinite sequence  $\<{\gr b_i}:i<\omega\>$

\ceq{\hfill {\mr b_i}}{\models}{p_{A,b\restriction i}({\gr z})\ \cup\ \{{\gr z}\notin^i\grD\}}

Let $n$ be the largest integer such that $\<{\gr b_i}:i<n\>$ satisfies the condition above. Then

\ceq{\hfill p_{A,b\restriction n}({\gr z})}{\imp}{{\gr z}\notin^n\grD}

and the lemma follows by compactness.
\end{proof}



\section{Notes and references}
\begin{biblist}[]\normalsize

\bib{CS}{article}{
   author={Chernikov, Artem},
   author={Simon, Pierre},
   title={Externally definable sets and dependent pairs},
   status={\href{https://arxiv.org/abs/1007.4468}{arXiv:1007.4468}},
   journal={Israel J. Math.},
   volume={194},
   date={2013},
   number={1},
   pages={409--425},
   %note={},
   %issn={0021-2172},
   %doi={10.1007/s11856-012-0061-9},
}



\end{biblist}


\end{document}


