% !TEX root = creche.tex
\chapter*{References}\label{references}
\addcontentsline{toc}{chapter}{References}
\vskip-3ex
\begin{biblist}[]\normalsize
\bib{BZ}{article}{
  author={Barbina, Silvia},
  author={Zambella, Domenico},
  title={A viewpoint on amalgamation classes},
  journal={Comment. Math. Univ. Carolin.},
  volume={51},
  date={2010},
  status={\href{https://arxiv.org/abs/1009.1789}{arXiv:1009.1789}},
  % number={4},
  % pages={681--691},
  % issn={0010-2628},
  % review={\MR{2858269}},
}\smallskip
\bib{Cameron}{article}{
  author={Cameron, Peter J.},
  title={The random graph},
  note={t.a. in The Mathematics of Paul Erdös III},
  status={\href{https://arxiv.org/abs/1301.7544}{arXiv:1301.7544}},
  %eprint={https://arxiv.org/abs/1301.7544},
  date={2013},
 }\smallskip
\bib{CS}{article}{
  author={Chernikov, Artem},
  author={Simon, Pierre},
  title={Externally definable sets and dependent pairs},
  status={\href{https://arxiv.org/abs/1007.4468}{arXiv:1007.4468}},
  journal={Israel J. Math.},
  volume={194},
  date={2013},
  number={1},
  pages={409--425},
  %note={},
  %issn={0021-2172},
  %doi={10.1007/s11856-012-0061-9},
}\smallskip
\bib{CZ}{article}{
  author={Colla, Eugenio},
  author={Zambella, Domenico},
  title={Ramsey's coheirs},
  date={2019},
  status={Submitted, \href{https://arxiv.org/abs/1901.04363}{arXiv:1901.04363}},
 }\smallskip
\bib{AK}{article}{
  author={Alex Kruckman},
  title={Counterexample to the omitting types in uncountable language},
  eprint={https://math.stackexchange.com/q/2434851},
  note={URL accessed 2019-02-28},
  %status={\href{https://math.stackexchange.com/q/2434851}},
  date={2017},
 }\smallskip 
\bib{Pelaez}{book}{
  author={Pel\'aez, Rodrigo},
  title={\href{http://www.ub.edu/modeltheory/documentos/ThesisRPP.pdf}{About the Lascar group}},
  series={PhD Thesis},
  publisher={Universitat de Barcelona, Departament de L\'ogica, Hist\'oria i Filosofia de la Ci\'encia},
  date={2008},
}\smallskip
\bib{Pillay}{book}{
  author={Pillay, Anand},
  title={Geometric stability theory},
  series={Oxford Logic Guides},
  volume={32},
  %note={Oxford Science Publications},
  %publisher={The Clarendon Press, Oxford University Press, New York},
  date={1996},
  % pages={x+361},
}\smallskip
\bib{Simon}{book}{
  author={Simon, Pierre},
  title={\href{http://www.normalesup.org/~simon/NIP_guide.pdf}{A guide to NIP theories}},
  series={Lecture Notes in Logic},
  volume={44},
  publisher={Association for Symbolic Logic, Cambridge University Press},
  date={2015},
%   pages={vii+156},
%   isbn={978-1-107-05775-3},
%   review={\MR{3560428}},
%   doi={10.1017/CBO9781107415133},
}\smallskip
\bib{TZ}{book}{
  author={Tent, Katrin},
  author={Ziegler, Martin},
  title={A course in model theory},
  series={Lecture Notes in Logic},
  volume={40},
  publisher={Association for Symbolic Logic, Cambridge University Press},
  date={2012},
  pages={x+248},
  %isbn={978-0-521-76324-0},
  %doi={10.1017/CBO9781139015417},
}\smallskip
\bib{Z16}{article}{
  author={Zambella, Domenico},
  title={On the diameter of Lascar strong types after Ludomir Newelski},
  conference={
    title={A tribute to Albert Visser},
  },
  book={
    % series={Tributes},
    % volume={30},
    publisher={Coll. Publ., [London]},
  },
  date={2016},
  status={\href{https://arxiv.org/abs/1605.00218}{arXiv:1605.00218}},
%  pages={231--236},
%  review={\MR{3559880}},
}\smallskip
\bib{Z15}{article}{
  author={Zambella, Domenico},
  title={Elementary classes of finite VC-dimension},
  journal={Arch. Math. Logic},
  volume={54},
  date={2015},
  status={\href{https://arxiv.org/abs/1412.5781}{arXiv:1412.5781}},
  %  number={5-6},
  %  pages={511--520},
  %  issn={0933-5846},
  %  review={\MR{3372605}},
  %  doi={10.1007/s00153-015-0424-0},
}\smallskip
\end{biblist}